---
title: "Predicting Country-Level Life Expectancy"
author: 'Team Last!: Parker Dingman, Ethan Donecoff, Karam Oubari, Pei Yi Zhuo'
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r load-packages-data}
library(tidyverse)
library(here)
library(rms)
life <- read_csv(here("data", "life_expectancy.csv"))
regions <- read_csv(here("data", "regions.csv"))
```

```{r add-region, results = "hide"}
regions <- regions %>%
  select(Country, Region) %>%
  mutate(Country = gsub("&", "and", Country))
initial_join <- life %>%
  left_join(regions, by = "Country") %>%
  select(Country, Region, Year:Schooling)
(na_countries <- initial_join %>%
  select(Country, Region) %>%
  distinct() %>%
  filter(is.na(Region)) %>%
  select(Country) %>%
  pull())
# Montenegro, Niue, and South Sudan do not appear in 'regions'.
# Alternate names of Cabo Verde, Czechia, and 
# Timor-Leste were obtained through Google.
regions <- regions %>%
  mutate(
    Country = case_when(
      Country == "Bahamas, The" ~ "Bahamas",
      Country == "Bolivia" ~ "Bolivia (Plurinational State of)",
      Country == "Brunei" ~ "Brunei Darussalam",
      Country == "Cote d'Ivoire" ~ "Côte d'Ivoire",
      Country == "Cape Verde" ~ "Cabo Verde",
      Country == "Central African Rep." ~ "Central African Republic",
      Country == "Congo, Repub. of the" ~ "Congo",
      Country == "Czech Republic" ~ "Czechia",
      Country == "Korea, North" ~ "Democratic People's Republic of Korea",
      Country == "Congo, Dem. Rep." ~ "Democratic Republic of the Congo",
      Country == "Gambia, The" ~ "Gambia",
      Country == "Iran" ~ "Iran (Islamic Republic of)",
      Country == "Laos" ~ "Lao People's Democratic Republic",
      Country == "Micronesia, Fed. St." ~ "Micronesia (Federated States of)",
      Country == "Burma" ~ "Myanmar",
      Country == "Korea, South" ~ "Republic of Korea",
      Country == "Moldova" ~ "Republic of Moldova",
      Country == "Russia" ~ "Russian Federation",
      Country == "Syria" ~ "Syrian Arab Republic",
      Country == "Macedonia" ~ "The former Yugoslav republic of Macedonia",
      Country == "East Timor" ~ "Timor-Leste",
      Country == "United Kingdom" ~ 
        "United Kingdom of Great Britain and Northern Ireland",
      Country == "Tanzania" ~ "United Republic of Tanzania",
      Country == "United States" ~ "United States of America",
      Country == "Venezuela" ~ "Venezuela (Bolivarian Republic of)",
      Country == "Vietnam" ~ "Viet Nam",
      TRUE ~ Country
    )
  )
life <- life %>%
  left_join(regions, by = "Country") %>%
  select(Country, Region, Year:Schooling)
# The regions of Montenegro, Niue, and South Sudan 
# were identified/confirmed through Google.
life$Region[life$Country == "Montenegro"] <- "EASTERN EUROPE"
# https://en.wikipedia.org/wiki/Category:Countries_in_Oceania
life$Region[life$Country == "Niue"] <- "OCEANIA"
life$Region[life$Country == "South Sudan"] <- "SUB-SAHARAN AFRICA"
rm(initial_join)
rm(na_countries)
rm(regions)
```

```{r filter-year, results = "hide"}
na_prop <- function(x) {
  x %>%
    is.na() %>%
    sum() %>%
    prod(1 / n()) 
}
years <- rep(list(NA), 16)
for(year in 2000:2015) {
  temp <- life %>%
    filter(Year == year) %>%
    summarise(across(everything(), na_prop)) %>%
    pivot_longer(cols = everything(),
                 names_to = "variable",
                 values_to = "na_prop") %>%
    pull(na_prop)
  names(temp) <- names(life)
  years[[year - 1999]] <- temp
}
names(years) <- 2000:2015
# 2011 is the latest year where no variable 
# has a missing rate greater than 0.22.
map_dbl(years, ~ any(`>`(., 0.22)))
# All the years where no variable has a missing rate greater 
# than 0.22 has a 'population' missing rate of 0.2185792.
map(years, ~ .[`>`(., 0.20)])
# pred_life will be used for predicting 2012 life expectancy
# using our 2011 model to test predictive power
pred_life <- life %>%
  filter(Year == 2012) %>%
  drop_na()
life <- filter(life, Year == 2011)
# Remove unnecessary objects
rm(na_prop)
rm(years)
rm(year)
rm(temp)
```

## Introduction and Data

### Research Question, Background, and Data

In this research project, we hope to most accurately predict life expectancy using economic, social, and health-related country level-data in conjunction with the regression analysis method of multiple linear regression. Therefore, our research question is: how can we most accurately predict a country's average life expectancy?

By telling us the average age of death in a population, life expectancy is a key metric for understanding a country's health. According to Max Roser, Esteban Ortiz-Ospina and Hannah Ritchie from Our World in Data, "Broader than the narrow metric of the infant and child mortality, which focus solely at mortality at a young age, life expectancy captures the mortality along the entire life course."[1] 

Over the course of history, life expectancy has risen dramatically. It is estimated that in pre-modern times, life expectancy worldwide was only about 30 years. Since the Industrial Revolution in the 18th and 19th centuries, many countries had large increases in life expectancy. Since the beginning of the 20th century, global average life expectancy has risen to about 70 years. However, there remain huge inequalities in this number. Currently (as of 2019), the Central African Republic has the lowest life expectancy of 53 years while Japan has the highest with 83 years.

In addition to the more obvious health-related connections to life expectancy, numerous pieces of academic literature have delved into the non-medical factors behind life expectancy. A major example is a longitudinal study conducted by Charles Lin, Eugene Rogot, Norman Johnson, Paul Sorlie, and Elizabeth Arias, which examined life expectancy by socioeconomic factors.[2] Academic literature such as this provides us with motivations to examine this topic on an international level, where we can look at various health-related and non-health-related factors that connect to life expectancy.

In terms of initial hypotheses of model selection, we expect that stronge predictors of life expectancy will be `Adult Mortality`, `infant deaths`, `under-five deaths`, and `Total expenditure`. We also predict that countries that have `Status` equal to "Developed" will have higher life expectancy than those that have `status` equal to "Developing".

Our primary data set is comprised of information that had been gleaned from the websites of the World Health Organization and the United Nations.[3] Each entry describes the health, social, and economic conditions for one of 193 countries in a given year from 2000 to 2015.[3] Because this data set lacks a variable that specifies the region of each country, we joined it with another data set that pairs country and region.[4] This secondary data set compiles information that can be found on the CIA World Factbook.[4] Both data were found on Kaggle and credit is given to Deeksha Russell, Duan Wang, Fernando Lasso, and the above organizations for contributing to the creation of our datasets.

\newpage

*Definitions of Relevant Variables:*

Response Variable:

- `Life expectancy`: Average life expectancy in years

Identifier Variable:

- `Country`: Name of country

Predictor Variables:

- `Region`: Area, subcontinent, or continent where a country is located

- `Adult Mortality`: Probability of dying between 15 and 60 years per 1000 population 
 
- `infant deaths`: Number of infant deaths per 1000 population

- `Hepatitis B`: Percentage of hepatitis B immunization coverage among 1-year-olds

- `Measles`: Number of reported measles cases per 1000 population

- `Total expenditure`: General government expenditure on health as a percentage of total government expenditure

- `Diphtheria`: Percentage of DTP3 immunization coverage among 1-year-olds

- `HIV/AIDS`: Deaths per 1000 live births HIV/AIDS (0-4 years)

- `Income composition of resources`: Human Development Index in terms of income composition of resources (index ranging from 0 to 1)


### Exploratory Data Analysis

First, we will look at some summary statistics of our response variable, `Life expectancy`.
Though it is possible to analyze the entire data set over all years, this creates difficulties 
with creating models. As one might expect, the average global life expectancy increased from
2000 to 2015. This relationship might make it unclear whether a rise in life expectancy is explained
by our predictor variables or if it is simply due to human development over time. As a result, we 
will only use data from one year to perform our analysis.

We will only analyze life expectancy for the year 2011 because it is the latest year among the years with the least missing data (no variable has a missing rate greater than 22%).

```{r summary-stats1, warning = FALSE}
life %>%
  filter(!is.na(`Life expectancy`)) %>%
  summarise(min = min(`Life expectancy`), 
            max = max(`Life expectancy`),
            range = max - min,
            mean = mean(`Life expectancy`), 
            median = median(`Life expectancy`), 
            Q1 = quantile(`Life expectancy`, .25), 
            Q3 = quantile(`Life expectancy`, .75), 
            iqr = IQR(`Life expectancy`), 
            sd = sd(`Life expectancy`)) %>%
  kable(digits = 3, caption = "Summary Statistics of Response Variable, Life Expectancy")
```

These summary statistics give a rough idea of the distribution of the response variable.
The median life expectancy (~73.3 yrs) is almost 3 years greater than the mean (~70.7 yrs).
Additionally, the median is closer to the third quartile than the first quartile. 
This suggests that life expectancy may be left-skewed, which we will evaluate further 
with visualizations.

Now, we will look at some summary visualizations of life expectancy, as well as other possibly relavant predictor variables.

```{r summary-viz, fig.height = 2, fig.width = 6}
p1 <- ggplot(data = life, mapping = aes(x = `Life expectancy`)) + 
  geom_histogram() + 
  labs(title = "Distribution of Life Expectancy (2011)",
       x = "Life Expectancy (Years)",
       y = "Count") + 
  theme(plot.title = element_text(size=11), axis.title=element_text(size=10))

p2 <- ggplot(data = life, mapping = aes(x = `Life expectancy`)) + 
  geom_density() + 
  labs(title = "Distribution of Life Expectancy (2011)",
       x = "Life Expectancy (Years)",
       y = "Proportion")

p3 <- ggplot(data = life, mapping = aes(x = `Life expectancy`, y = "")) + 
  geom_boxplot() + 
  labs(title = "Distribution of Life Expectancy (2011)",
       x = "Life Expectancy (Years)",
       y = "") + 
  theme(plot.title = element_text(size=11), axis.title=element_text(size=10))

by_status <- life %>%
  filter(!is.na(`Life expectancy`)) %>%
  group_by(`Status`) %>%
  summarise(mean = mean(`Life expectancy`))

(p1 + p3)
```

```{r, fig.height = 2, fig.width = 6}
p5 <- ggplot(data = by_status, aes(x = `Status`, y = `mean`)) +
  geom_bar(stat = "identity") +
  labs(x = "Status",
       y = "Life Expectancy",
       title = "Life Expectancy",
       subtitle = "by country status") + 
  theme(plot.title = element_text(size=11), 
        axis.title=element_text(size=10),
        plot.subtitle = element_text(size=9))
```

```{r, fig.height = 3, fig.width = 6}
p4 <- ggplot(data = life, aes(x = `Adult Mortality`, 
                              y = `Life expectancy`)) +
  geom_point(aes(size = `infant deaths`)) +
  labs(x = "Adult Mortality",
       y = "Life Expectancy",
       size = "Infant deaths\nper 1000",
       title = "Life Expectancy vs. Adult Mortality") + 
  theme(plot.title = element_text(size=11), 
        axis.title=element_text(size=10),
        legend.title=element_text(size=10),
        legend.text=element_text(size=9))

(p4 + p5)
```

In terms of the response variable, we see that `Life expectancy` is skewed to the left with a center just over 70 years. `Life expectancy` ranges from about 50 years to 90 years. 

For relationships with predictor variables, we see that a strong negative relationship exists between `Life expectancy` and `Adult Mortality`. We also a see a strange diversion in the points between countries that have lower and higher values for `Adult mortality`. There appears to be two groups of countries in the scatter plot, a smaller group that exhibits a greater `Life expectancy` penalty for a given increase in `Adult Mortality` than the larger group. We thought that this is because some of those countries have higher infant mortalities, driving life expectancy down. However, when points are sized according to `infant deaths`, this is shown not to be the case. We also see much higher `Life expectancy` among countries with the `status` "developed" than "developing", with a difference of about 12 years.


```{r world map}
library(ggplot2)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(rgeos)

world <- ne_countries(scale = "medium", returnclass = "sf")

map_data <- life %>%
  mutate(Country = case_when(
         Country == "United States of America" ~ "USA", 
         Country == "Russian Federation" ~ "Russia", 
         TRUE ~ Country)) %>%
  select(Country, `Life expectancy`)

map <- map_data("world")

life_map <- left_join(map, map_data, by = c("region" = "Country"))

ggplot(data = life_map, mapping = aes(long, lat, group = group)) +
  coord_fixed(1) +
  geom_polygon(
    # color = "black",
    aes(fill = as.numeric(`Life expectancy`))
  ) +
  scale_fill_gradient(low = "red", high = "green", 
                      name = "2011 Life Expectancy in Years") +
  labs(title = "2011 Life Expectancy Across The World", 
       x = "Longitude", y = "Latitude")

#https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html
```


Above we can see the differences in life expectancy around the world. This map allows us to visualize how life expectancy varies between regions. It appears that that the Sub-Saharan African region has the lowest life expectancy, while regions like North America and Western Europe have the highest life expectancy.

```{r correlated-predictors, fig.height = 3, fig.width = 8}
infant_five <- ggplot(data = life, aes(x = `infant deaths`, 
                                       y = `under-five deaths`)) +
  geom_point() +
  labs(x = "Infant Deaths",
       y = "Under-five Deaths")

schooling_income <- ggplot(data = life, 
                           aes(x = Schooling, 
                               y = `Income composition of resources`)) +
  geom_point()

title <- plot_annotation("Correlated Predictors")

infant_five + schooling_income + title
```

These two scatter plots demonstrate that the predictor variable `infant deaths` is correlated with the predictor variable `under-five deaths`. Likewise, two other predictor variables, `Income composition of resources` and `Schooling`, are clearly associated with one another. This finding will be addressed as we construct our model.

## Methodology

In order to answer the research question, we will be using multiple linear regression techniques due to the quantitative nature of our response variable, `Life expectancy`.

In order to create a model capable of predicting life expectancy, we will use model selection to cull the number of potential predictor variables. We will then assess whether certain interaction terms should be added to the model before addressing multicollinearity. Finally, with the set of predictor variables finalized, we ascertain whether any variable transformations are necessary through condition checking and identifying outliers by examining model diagnostics. 

### Model Selection

The goal of our analysis is to calculate the most precise prediction of the response variable (`Life expectancy`). To do this we will perform forward and backward selection using AIC and BIC. Forward selection entails adding variables recursively using AIC or BIC as the criterion while backward selection means dropping variables one at a time, starting with our full model (Appendix Table 9), that are deemed irrelevant based on AIC or BIC.

```{r full-model}
life <- life %>%
  drop_na()
full_model <- lm(`Life expectancy` ~ Region + Status + `Adult Mortality` + 
                   `infant deaths` + Alcohol + `percentage expenditure` + 
                   `Hepatitis B` + Measles + BMI + `under-five deaths` + 
                   Polio + `Total expenditure` + Diphtheria + `HIV/AIDS` + 
                   GDP + Population + `thinness  1-19 years` + 
                   `thinness 5-9 years` + `Income composition of resources` + 
                   Schooling, data = life)
```

```{r model-selection, results = "hide"}
# Backward selection
backward_aic <- step(full_model, direction = "backward")
backward_bic <- step(full_model, direction = "backward", k = log(nrow(life)))
# Forward selection
int_model <- lm(`Life expectancy` ~ 1, data = life)
forward_aic <- step(int_model, formula(full_model), direction = "forward")
forward_bic <- step(int_model, formula(full_model), 
                    direction = "forward", k = log(nrow(life)))
```

```{r model-comparison}
models <- list(backward_aic, backward_bic, 
               forward_aic, forward_bic)
methods <- c("Backward (AIC)", "Backward (BIC)",
             "Forward (AIC)", "Forward (BIC)")

out <- tibble(
  `Selection Method` = NA,
  `AIC` = NA,
  `BIC` = NA,
  `Adjusted R-squared` = NA
)
for(i in 1:length(models)) {
  temp <- models[[i]] %>%
    glance() %>%
    select(AIC, BIC, `Adjusted R-squared` = adj.r.squared) %>%
    mutate(`Selection Method` = methods[[i]])
  out <- rbind(out, temp)
}

kable(out[-1, ], caption = "Potential Models", digits = 3)

# glance(backward_aic) %>% 
#   select(AIC, BIC, adj.r.squared) #this will become the selected model
# glance(backward_bic) %>% 
#   select(AIC, BIC, adj.r.squared) 
# glance(forward_aic) %>% 
#   select(AIC, BIC, adj.r.squared)
# glance(forward_bic) %>% 
#   select(AIC, BIC, adj.r.squared)
selected_model <- backward_aic
```

Based on AIC and adjusted $R^2$, we prefer the model we found through backward selection using AIC. This model results in the highest adjusted $R^2$ out of all four models. Moreover, this model is the only model out of the four that is superior to the remaining three models in more than one metric (AIC and adjusted $R^2$). 

### Potential Interaction Terms

Before proceeding with the model, we will look at some potential interaction terms
to see if they should be added to the selected model. We will begin by looking 
at the interaction between Region and HIV/AIDS. This is because we might suspect
that the impact of HIV/AIDS on life expectancy changes by region due to 
disparate access to relatively new treatments. The interaction between Region 
and Income composition of resources is also shown. The table below shows the 
selected model including these interaction terms, but only displays the interaction 
terms. 

```{r interactions}
interaction_model <- lm(`Life expectancy` ~ Region + `Adult Mortality` + 
                   `infant deaths` + Region*`HIV/AIDS` + 
                     Region*`Income composition of resources` +
                   `Hepatitis B` + Measles + `under-five deaths`+ `Total expenditure` + 
                     Diphtheria + `HIV/AIDS`+ `Income composition of resources` + 
                   Schooling, data = life)

interaction_model <- tidy(interaction_model) %>%
  filter(grepl(":", term))

kable(interaction_model, digits = 3, 
      caption = "Selected Model Interaction Terms")
```

As shown in the table above, some of the groups do not have enough data to generate model coefficients. For the model coefficients calculated, there are no significant interaction terms for the variables investigated. Thus, there is not enough evidence to say that
there is a significant interaction term between Region and HIV/AIDS or Region 
and Income composition of resources. 

\newpage

### Multicollinearity

```{r multicollinearity}
vif(selected_model) %>% 
  tidy() %>% 
  rename(Variable = names, VIF = x) %>%
  kable(caption = "Predictor Variable VIFs", digits = 3)
```
Variables with a VIF > 10 will have issues with multicollinearity. `infant deaths` and `under-five deaths` are clearly highly correlated (this makes a lot of sense in the context of the data). `Income composition of resources` appears to be correlated with `Schooling`. These two relationships were visualized in the EDA. 

We will try models without `infant deaths` or without `under-five deaths` and then use model comparison techniques to decide on which of these two variables should be removed. Likewise, we will compare models with only `Income composition of resources` or `Schooling` and keep only one of the two. 

```{r infant-five}
inf <- lm(`Life expectancy` ~ Region + `Adult Mortality` +
            `infant deaths` + `Hepatitis B` + `Measles` + 
            `Total expenditure` + `Diphtheria` + `HIV/AIDS` +
            `Income composition of resources` + `Schooling`,
          data = life)
five <- lm(`Life expectancy` ~ Region + `Adult Mortality` +
             `Hepatitis B` + `Measles` + `under-five deaths` +
             `Total expenditure` + `Diphtheria` + `HIV/AIDS` +
             `Income composition of resources` + `Schooling`,
           data = life)
model_with <- function(variable, model) {
  model %>%
    glance() %>%
    select(AIC, BIC, adj.r.squared) %>%
    mutate(`Included Variable` = variable, .before = AIC)
}
model_with("Infant Deaths", inf) %>%
  rbind(model_with("Under-five Deaths", five)) %>%
  kable(caption = "Infant Deaths vs. Under-five Deaths",
        digits = 4)
```

When comparing two models that are identical except that one includes `under-five deaths` and the other includes `infant deaths`, the one that includes `infant deaths` has lower values of AIC and BIC as well as a higher value of adjusted $R^2$.

```{r income-schooling}
sch <- lm(`Life expectancy` ~ Region + `Adult Mortality` +
            `infant deaths` + `Hepatitis B` + `Measles` + 
            `Total expenditure` + `Diphtheria` + `HIV/AIDS` + 
            `Schooling`, data = life)
inc <- lm(`Life expectancy` ~ Region + `Adult Mortality` +
            `infant deaths` + `Hepatitis B` + `Measles` + 
            `Total expenditure` + `Diphtheria` + `HIV/AIDS` +
            `Income composition of resources`, data = life)
model_with("Schooling", sch) %>%
  rbind(model_with("Income Composition of Resources", inc)) %>%
  kable(caption = "Schooling vs. Income Composition of Resources",
        digits = 3)
selected_model <- inc
```

\newpage

Likewise, given two models that have the same predictors except that one includes `Schooling` while the other includes `Income composition of resources`, the one that includes `Income composition of resources` is superior in terms of AIC, BIC, and adjusted $R^2$.

### Model Conditions and Diagnostics

```{r augment}
selected_model_aug <- augment(selected_model)
selected_model_aug <- selected_model_aug %>%
  mutate(obs_num = 1:nrow(selected_model_aug))
```

```{r normality-viz, fig.height = 3, fig.width = 8}
resid_hist <- selected_model_aug %>%
  ggplot(aes(x = .std.resid)) +
  geom_histogram() +
  labs(title = "Distribution of Standardized Residuals",
       x = "Residual",
       y = "Count")
resid_qq <- selected_model_aug %>%
  ggplot(aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "QQ Plot of Standardized Residuals",
       x = "Theoretical",
       y = "Actual")
resid_hist + resid_qq
```

Judging from the histogram and QQ plot above, the standardized residuals appear to be normally distributed, thus fulfilling the normality condition. 

```{r residual-plot, fig.height = 3, fig.width = 6}
ggplot(data = selected_model_aug, aes(x = .fitted, y = .std.resid)) +
  geom_point(alpha = 0.7) + 
  geom_hline(yintercept = 0, color = "red") +
  geom_hline(yintercept = -2, color = "black", linetype = "dotted") +
  geom_hline(yintercept = 2, color = "black", linetype = "dotted") +
  geom_hline(yintercept = -3, color = "red", linetype = "dotted") +
  geom_hline(yintercept = 3, color = "red", linetype = "dotted") +
  labs(x = "Predicted", y = "Standardized Residuals",
       title = "Standardized Residuals vs. Predicted")
```

```{r response-outliers, results = "show"}
high_res <- selected_model_aug %>%
  filter(.std.resid > 2 | .std.resid < -2) %>%
  left_join(life) %>%
  select(Country, `Standardized Residual` = .std.resid) 
kable(high_res, caption = "Life Expectancy Outliers", digits = 3)
```

The plot above depicts standardized residuals on the y-axis and predicted values on the x-axis. We see no major departure from the constant variance assumption. However, the two countries Iraq and Lesotho are moderate outliers. Two other countries, Angola and Sierra Leone, are severe outliers in terms of their life expectancy. Specifically, Iraq has a much higher life expectancy than what we would expect given the values of its predictor variables while Angola, Lesotho, and Sierra Leone have far lower values for life expectancy than what was predicted. 

Moreover, the standardized residuals do not display a discernible pattern when plotted against any of the quantitative variables. The visualization illustrating this observation can be found below, in the appendix section. This combined with the similar lack of any relationship in the plot of standardized residuals vs. predicted values above indicates that the linearity assumption holds. 

```{r region, results = "hide"}
count(life, Region)
```

Lastly, we do not believe the independence condition for multiple linear regression can be considered fulfilled in our case. This is because countries are likely associated with one another in ways that we cannot capture with our model even though we did include `Region` as a predictor variable to try to address spatial correlation. For instance, some categories of `Region` are especially large (`SUB-SAHARAN AFRICA` encompasses 49 nations), yet others are quite small (`BALTICS` includes only 3 nations). Those countries within `SUB-SAHARAN AFRICA` may have yet more relationships with one another for which we simply cannot account with our current data set.

```{r threshold}
# (2(p+1))/n
leverage_threshold <- 2 * (20 + 1) / nrow(selected_model_aug)
# kable(leverage_threshold,
#       caption = "Leverage Threshold:",
#       align = "l")
```

```{r high-leverage}
high_lev <- selected_model_aug %>%
  filter(.hat > leverage_threshold) %>%
  left_join(life) %>%
  select(Country, Leverage = .hat) 
kable(high_lev, caption = "Predictor Variable Outliers", digits = 3)
```

```{r influential, results = "hide"}
selected_model_aug %>%
  filter(.cooksd > .5) %>%
  left_join(life) %>%
  select(Country, .cooksd) 
```

In terms of leverage and influential points, the 12 countries in the table above have high leverage (leverage > leverage_threshold = 0.3230769) while no countries are influential (Cook's Distance > 0.5). 

## Results

### Final Model

```{r selected-model}
(table_selected_model <- kable(tidy(selected_model),
                               caption = "Selected Model",
                               digits = 3))

r_sq <- selected_model %>%
  glance() %>%
  select(r.squared) %>%
  prod(100) %>%
  round(2)

aic <- selected_model %>%
  glance() %>%
  select(AIC) %>%
  round(2)

bic <- selected_model %>%
  glance() %>%
  select(BIC) %>%
  round(2)

am_coef <- selected_model$coefficients["`Adult Mortality`"] %>%
  prod(100) %>%
  round(2)
```

This is our final model. Around `r r_sq`% of the variation in the `Life expectancy` is explained by the regression model above, which contains the predictor variables: `Region`, `Adult Mortality`, `infant deaths`, `Hepatitis B`, `Measles`, `Total expenditure`, `Diphtheria`, `HIV/AIDS`, and `Income composition of resources`. The AIC is about `r aic`, and the BIC is approximately `r bic`. 


### Key Findings and Conclusions

One interesting key finding when looking at the variables included in our selected model is that the predictors for life expectancy were extremely multifaceted. Variables that related to health-related, economic, and social measures of a country all play an important part in predicting life expectancy.

Something else that stands out is the extremely high estimated coefficient (31.747) of `Income composition of resources`. However, this makes sense when it is understood that the variable's unit (HDI) is measured on a scale of 0 to 1. Therefore, it is more informative to interpret this variable in smaller units than an increase by one, such as an increase by 0.01. In that case, for each additional 0.01 increase in HDI, we expect life expectancy to increase by about 0.32 years or about 4 months, on average, holding all other predictor variables constant. 

Additionally, when looking at the selected model, a few other key findings stand out. First, it was interesting to see the estimated coefficients of the different regions, which can be interpreted as the expected average life expectancy above the baseline of Asia (excluding Near East), holding all other predictors constant. The region of Sub-Saharan Africa had the lowest coefficient of -2.576 while the region Western Europe had the highest coefficient of 4.905. According to our model, these discrepancies are solely because of the difference in geographic region, which logically doesn't tell us the full story. Therefore, it's reasonable to assume that some information that differentiates geographic regions was missing from our data. If this project were to be expanded, it would be important to include more than 23 potential variables in the full model.

Also, it is strange to see that according to our model, an increase in infant deaths is expected to increase life expectancy, on average (due to the positive estimated coefficient). This clearly doesn't make much sense. However, this can be taken with a grain of salt since the p-value is relatively high at 0.409. This means that assuming the null hypothesis that this variable's coefficient is equal to 0 is true, the probability of observing a coefficient at least as extreme as the one we've observed is not unlikely.

In terms of our initial hypotheses, the coefficient of only one variable that we thought would have a major effect on `Life expectancy` is significant: `Adult Mortality`. Per 100 adults (ages 15 to 60) out of 1000 that die, life expectancy is expected to decrease by around `r -1 * am_coef` years, holding all else constant. The other variables that we hypothesized at the beginning of the project, `infant deaths`, `under-five deaths`, and `Total expenditure` either are not included in the final model or are not significantly different from 0 at the 0.05 significance level. Other significant variables are `Income composition of resources` and `HIV/AIDS` with p-values near 0. 

\newpage

### Assessing Predictive Power

To evaluate the predictive power of our model, we predicted the life expectancy
for each country during 2012 and compared to the observed values from the 
original `life` dataset. We chose 2012 because we thought that one important application of a predictive model for life expectancy would be to predict life expectancy for the following year. The residual plot is shown below.

```{r prediction, fig.height = 4, fig.width = 8}
# Dataframe of baseline observations for predicting
base_obs <- life %>%
  select(Country, Region, `Adult Mortality`, `infant deaths`, `Hepatitis B`, Measles,
         `Total expenditure`, Diphtheria, `HIV/AIDS`, `Income composition of resources`) %>%
  # Remove Eritrea since it has NAs for 2012
  filter(Country != "Eritrea")
  
# Predict 2012 life expectancy
pred_life <- pred_life %>%
  mutate(pred_LifeExp = predict(selected_model, base_obs, na.action = na.exclude))

# Calculate residuals 
pred_life <- pred_life %>%
  mutate(res = `Life expectancy` - pred_LifeExp, 
         tss = (`Life expectancy` - mean(`Life expectancy`))^2,
         rss = (res)^2)

# R-squared
rsq <- 1 - sum(pred_life$rss)/sum(pred_life$tss)

# Visualize residuals
ggplot(data = pred_life, aes(x = `Life expectancy`, y = res)) +
  geom_point(alpha = 0.7) + 
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "Predicted 2012 Life Expectancy", y = "Residuals",
       title = "Residuals vs. Predicted",
       subtitle = "Predicted 2012 Life Expectancy")
```

As seen above, the residuals for 2012 show no discernable pattern and display a 
relatively constant vertical spread. The residuals are positive on average, 
with a sum of `r sum(pred_life$res)`. This indicates that the model slightly
underpredicts life expectancy for the next year (2012), but appears fairly accurate
overall. We also see that $R^2=$ `r rsq`. This means that over 90% of the 
variability in 2012 Life expectancy can be explained by our model, further 
demonstrating that it can predict future life expectancy relatively accurately.

## Discussion + Conclusion

As mentioned in the Model Conditions and Diagnostics section, our model struggles when it comes to the independence condition. Indeed, the finding in the EDA that there are two distinct relationships between `Life expectancy` and `Adult mortality` among the countries seems to suggest that there is some unseen way in which countries are systematically related to each other. 

```{r no-response-outliers, results = "hide"}
table_selected_model

life %>% 
  filter(!(Country %in% high_res$Country)) %>%
  lm(`Life expectancy` ~ Region + `Adult Mortality` + `infant deaths` +
       `Hepatitis B` + Measles + `Total expenditure` + Diphtheria +
       `HIV/AIDS` + `Income composition of resources`, data = .) %>%
  tidy() %>%
  kable(caption = "Model Without Angola, Iraq, Lesotho, and Sierra Leone", 
        digits = 3)

model_without <- function(country) {
  life %>% 
  filter(Country != country) %>%
  lm(`Life expectancy` ~ Region + `Adult Mortality` + `infant deaths` +
       `Hepatitis B` + Measles + `Total expenditure` + Diphtheria +
       `HIV/AIDS` + `Income composition of resources`, data = .) %>%
  tidy() %>%
  kable(digits = 3, caption = paste("Model Without", country)) 
}

map(high_res$Country, model_without)
```

```{r no-high-leverage, results = "hide"}
table_selected_model

life %>% 
  filter(!(Country %in% high_lev$Country)) %>%
  lm(`Life expectancy` ~ Region + `Adult Mortality` + `infant deaths` +
       `Hepatitis B` + Measles + `Total expenditure` + Diphtheria +
       `HIV/AIDS` + `Income composition of resources`, data = .) %>%
  tidy() %>%
  kable(caption = "Model Without High Leverage Countries", digits = 3)
```

Another caveat for our model is that the data set used to fit the model contains outliers that have an outsized impact on the model itself. If we remove the four countries with large-magnitude standardized residuals from the data set and refit the model, we observe a dramatic effect on the p-value of `Total expenditure`. Whereas the coefficient had been positive and non-significant with a p-value above 0.2 in the model with the four countries, it becomes very significant with a p-value of around 0.008 once these observations are absent from the data set. Upon closer inspection, we found that individually removing Iraq, Lesotho, and Sierra Leone reduces the p-value of `Total expenditure`. However, removing Angola increases the p-value. Thus, it appears that it is the former three countries that are behind this effect. On the other hand, there are no differences between the model that excludes high leverage observations and the selected model that are as dramatic as the one between the model that excludes large-magnitude standardized residual observations and the selected model.

\newpage

## Appendix

```{r disp-full-model}
tidy(full_model) %>%
  kable(caption = "Full Model With All Possible Predictor Variables", 
        digits = 3)
```

```{r linearity-viz, fig.height = 10, fig.width = 8}
# https://www.youtube.com/watch?v=nqI5CmdUYRQ
plot_predictor <- function(variable) {
  ggplot(data = selected_model_aug, 
         aes(x = .data[[variable]], 
             y = .std.resid)) +
    geom_point() +
    labs(x = variable,
         y = "Standardized Residual")
}
plots <- map(names(selected_model_aug)[3:10], plot_predictor)
title <- plot_annotation("Checking Linearity")
(plots[[1]] + plots[[2]] + plots[[3]] + plots[[4]]) /
  (plots[[5]] + plots[[6]] + plots[[7]] + plots[[8]]) + title
```

\newpage

## References

[1] https://ourworldindata.org/life-expectancy 

[2] https://europepmc.org/article/med/12785422/reload=0#impact 

[3] https://www.kaggle.com/kumarajarshi/life-expectancy-who. 

[4] https://www.kaggle.com/fernandol/countries-of-the-world
