---
title: "Topic ideas"
author: "Team Last!: Parker Dingman, Ethan Donecoff, Karam Oubari, Pei Yi Zhuo"
date: "March 4, 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```


```{r load-packages-data}
# Load packages
library(tidyverse)
library(here)

# Load data
flying <- read_csv(here("data", "flying.csv"))
life <- read_csv("../data/life_expectancy.csv")
```

Put all data sets in the `data` folder. 

## Data Set 1 (Flying)

### Description 

### Data dictionary 

### `glimpse` of data

```{r glimpse-flying}
glimpse(flying)
```

## Data Set 2 (Life Expectancy)

### Description 

Data-set related to life expectancy, health factors, and economic data for 193 countries collected from WHO data repository and from United Nation.

### Data dictionary 

- Country
- Year
- Status (Developed or Developing status)
- Life expectancy (Life Expectancy in age)
- Adult Mortality (Adult Mortality Rates of both sexes (probability of dying between 15 and 60 years per 1000 population))
- infant deaths (Number of Infant Deaths per 1000 population)
- Alcohol (Alcohol, recorded per capita (15+) consumption (in litres of pure alcohol))
- percentage expenditure (Expenditure on health as a percentage of Gross Domestic Product per capita(%))
- Hepatitis B (Hepatitis B (HepB) immunization coverage among 1-year-olds (%))
- Measles (Measles - number of reported cases per 1000 population)
- BMI (Average Body Mass Index of entire population)
- under-five deaths (Number of under-five deaths per 1000 population)
- Polio (Polio (Pol3) immunization coverage among 1-year-olds (%))
- Total expenditure (General government expenditure on health as a percentage of total government expenditure (%))
- Diphtheria (Diphtheria tetanus toxoid and pertussis (DTP3) immunization coverage among 1-year-olds (%))
- HIV/AIDS (Deaths per 1 000 live births HIV/AIDS (0-4 years))
- GDP (Gross Domestic Product per capita (in USD))
- Population (Population of the country)
- thinness  1-19 years (Prevalence of thinness among children and adolescents for Age 10 to 19 (%))
- thinness 5-9 years (Prevalence of thinness among children for Age 5 to 9(%))
- Income composition of resources (Human Development Index in terms of income composition of resources (index ranging from 0 to 1))
- Schooling (Number of years of Schooling (years))

### `glimpse` of data

```{r glimpse-life}
glimpse(life)
```


## Data Set 3 (optional)

### Description 

### Data dictionary 

### `glimpse` of data



