---
title: "Topic ideas"
author: "Team Name: Team member 1, Team member 2, Team member 3, Team member 4"
date: "Update date"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```


```{r load-packages-data}
# Load packages
library(tidyverse)
library(here)

# Load data
flying <- read_csv(here("data", "flying.csv"))
life <- read_csv("../data/life_expectancy.csv")
```

Put all data sets in the `data` folder. 

## Data Set 1 (Flying)

### Description 

### Data dictionary 

### `glimpse` of data

```{r glimpse-flying}
glimpse(flying)
```

## Data Set 2 (Life Expectancy)

### Description 

Data-set related to life expectancy, health factors, and economic data for 193 countries collected from WHO data repository and from United Nation.

### Data dictionary 

### `glimpse` of data

```{r glimpse-life}
glimpse(life)
```


## Data Set 3 (optional)

### Description 

### Data dictionary 

### `glimpse` of data



